Thermodynamic fragments for magic states in quantum computation

### 2021-03-25 magic_paper.tex
Resolved clashes in edits.

\newpage
\section{Preliminaries and definitions}
\label{sec:ps}

\subsection{Stabilizer Theory}\label{sec:so}

Let $\{\ket{k}\}_{k \in \mathbb{Z}_d}$ be the standard computational basis, defined over $\mathbb{Z}_d = \{ 0, 1, \dots,d-1 \}$, with $d$ an odd prime. 
Since the field has character $d$, addition and multiplication on the field are always considered modulo $d$.
The Hilbert space of any system associated with this scheme is $\cal{H}_d \coloneqq \spanv{\ket{k}: k \in \mathbb{Z}_d}$.

The generalised Pauli matrices $X, Z$ can be defined by their respective roles as shift and phase operators,
\begin{align}
    X \ket{k} &= \ket{k + 1} \label{eq:xpauli}\\
	Z \ket{k} &= \omega^k \ket{k}, \label{eq:zpauli}
\end{align}
where $\omega \coloneqq e^{2\pi i/d}$ is the $d$-th root of unity and addition is modulo $d$. The Hilbert space $\cal{H}_d$ is associated with a phase space $\cal{P}_d = \mathbb{Z}_d \times \mathbb{Z}_d$, where every point $\bmx \coloneqq (x, p)$ corresponds to a displacement operator, defined as
\begin{equation}\label{eq:ddef}
    D_{\bmx} \coloneqq \tau^{x p} X^{x} Z^{p},\ (x, p) \in \cal{P}_d.
\end{equation}
The phase factor $\tau \coloneqq -\omega^{1/2}$ ensures unitarity.
For a composite system with product dimension $d = d_1 \dots d_n$ and Hilbert space, $\cal{H}_d = \cal{H}_{d_1} \otimes \dots \otimes \cal{H}_{d_n}$, the displacement operators are defined as
\begin{equation}\label{eq:composited}
    D_{\bmx} \coloneqq D_{(x_1, p_1)} \otimes \dots \otimes D_{(x_n, p_n)},
\end{equation}
where 
\begin{align*}
	\bmx \coloneqq (x_1, p_1, x_2, p_2, \dots, x_n, p_n) \in \cal{P}_{d_1} \times \dots \times \cal{P}_{d_n} \eqqcolon  \cal{P}_d,
\end{align*}
denotes the phase space point for an $n$--copy system.

The displacement operators form the Heisenberg-Weyl group~\cite{Folland_1989, Bengtsson_2006} under matrix multiplication modulo phases,
\begin{equation}\label{eq:gp}
    {\rm{HW}}_d^n \coloneqq \{ \tau^k D_{\bmx}: k \in \mathbb{Z}_d, \bmx \in \cal{P}_d^n\}.
\end{equation}
The Clifford operations $ \cal{C}_d^n $ are then defined as the set of unitaries that normalise the Heisenberg-Weyl group. We may define the pure stabiliser states as those states obtained by acting on $|0\>$ with Clifford unitaries. Finally, we define $\stab$ as the convex hull of all pure stabiliser states, namely all probabilistic mixtures of states of the form $U|0\>\<0|U^\dagger$ where $U$ is some Clifford unitary. The stabiliser theory forms a magic resource theory $\R = (\stab, \so)$
and the free operations are stabiliser operations ($\so$), defined as compositions of preparation of computational basis states, Clifford unitaries, measurement in the computational basis, and the ability to discard subsystems.

@@

We define the phase-point operators at every phase space point $\bmx \in \cal{P}_d$,
\begin{align}
	A_{\bmx} \coloneqq \frac{1}{d} \sum_{\bmz \in \cal{P}_d} \omega^{\eta(\bmx, \bmz)} D_{\bmz}, \label{eq:ax}
\end{align}
where $\eta(\bmx, \bmz)$ is the standard symplectic form, given explicitly by
\begin{equation}
	\eta(\bmx, \bmz) \coloneqq \bmz^T \begin{pmatrix}
		0  & \id \\ %\mathbbm{O}_n  & \id_n \\
		-\id & 0 \\ %-\id_n & \mathbbm{O}_n \\
	\end{pmatrix} \bmx,
\end{equation}
where $0, \id$ denote the $n\times n$ zero and identity matrices.

The phase-point operators form an orthogonal Hermitian operator basis as shown in~\cref{app:wigner}.
Therefore, any quantum state $\rho \in \cal{B}(\cal{H}_d)$ can be expressed as a linear combination of them,
\begin{equation}
    \rho = \sum_{\bmz \in \cal{P}_d} \W[\bmz]{\rho} A_{\bmz},
\end{equation}
where the coefficient vector $\W{\rho}$ is the Wigner distribution of state $\rho$,
\begin{equation}\label{eq:wstate}
    \W[\bmx]{\rho} \coloneqq \frac{1}{d}\tr[A_{\bmx} \rho].
\end{equation}

For any quantum state $\rho$, the Wigner distribution $W_\rho(\bmx)$ is readily seen to be a $d^2$--dimensional quasi-probability distribution over $\cal{P}_d$ (see~\cref{app:wigner} for details).  In~\cref{fig:wstate_examples}, we show Wigner distributions of different types of qutrit states.
\begin{figure}%
    \centering
    \subfigure[][]{%
    \label{fig:maxmix}%
    \includegraphics[height=2cm]{figs/maxmixed.pdf}
    %\caption{Maximally mixed state $\frac{1}{3}\id$}%
    }\hspace{8pt}%
    \subfigure[][]{%
    \label{fig:zero}%
    \includegraphics[height=2cm]{figs/zerostate.pdf}
    %\caption{Zero state $\ketbra{0}{0}$}%
    }\\
    \subfigure[][]{%
    \label{fig:bound}%
    \includegraphics[height=2cm]{figs/boundstate.pdf}
    %\caption{Bound state}%
    }\hspace{8pt}%
    \subfigure[][]{%
    \label{fig:strange}%
    \includegraphics[height=2cm]{figs/strangestate.pdf}
    %\caption{Strange state $\ketbra{S}{S}$}%
    }
    \caption{\textbf{Qutrit Wigner distributions of varying magic.} 
    \subref{fig:maxmix} Maximally mixed state $\frac{1}{3}\id$; \subref{fig:zero} Stabilizer zero state $\ketbra{0}{0}$; \subref{fig:bound} A non-stabiliser Wigner-positive state; \subref{fig:strange} Magic Strange state $\ket{{\rm{S}}} = \frac{1}{\sqrt{2}}(\ket{1} - \ket{2})$, coined in~\cite{cit:veitch2}.
    }%
    \label{fig:wstate_examples}
\end{figure}

The specific form of~\cref{eq:woperation} is chosen so that Wigner distributions of operations act as transition matrices for Wigner distributions of states, $\W{\E(\rho)} = \W{\E}\W{\rho}$.
In particular, $\cptp$ operations that map between density operators of equal dimensions and have non-negative Wigner distributions correspond to stochastic matrices, as shown in~\cref{app:wigner}

\subsection{Magic monotones}
\label{sec:mono}

We highlight the defining property of a monotone, as it is useful in later discussion.
\begin{definition}[\textbf{Resource monotone}]\label{def:mono}
    Let $\R = (\F, \O)$ be a resource theory.
    A resource monotone $\M$ is a projection from the set of quantum states of the theory onto the real line, so that $\M$ is monotonically decreasing under free operations,
    \begin{equation}
        \M(\rho_1) \leq \M(\rho_2)\ \text{whenever}\ \rho_1 \prec_{\R} \rho_2.
    \end{equation}
\end{definition}

One of the most fundamental and commonly used magic monotones is the \emph{mana} of a state~\cite{cit:veitch2}, defined as
\begin{equation}
    \mana{\rho} \coloneqq \ln{(2\hspace{1pt}\sn{\rho}+1)},
\end{equation}
where the \emph{sum-negativity}~\cite{cit:veitch2} is the sum of the negative components in $\W{\rho}$,
\begin{equation}
    \sn{\rho} \coloneqq \sum\limits_{\bmx: \W[\bmx]{\rho} < 0} \abs{\W[\bmx]{\rho}}.
\end{equation}
Mana is an additive\footnote{It satisfies the condition $\mana{\rho_1 \otimes \rho_2} = \mana{\rho_1} + \mana{\rho_2}$ which is practical in distillation settings.} magic monotone, so it provides an analytical, necessary condition for many-copy magic state interconvertibility.
The magic monotone of \emph{max--thauma}~\cite{Wang_2018} has also been introduced more recently for qudits of odd prime dimension.

@@

In~\cref{fig:lcs}, we illustrate the Lorenz curve $\lc{\rho_{\rm{S}}}{\sigma}$\footnote{Hereinafter, we may omit obvious variable dependencies like $\epsilon$ and $n$ for clarity} of pure and noisy $n$--copy Strange states in some thermal $\sigma$--fragment.
Due to~\cref{lem:lcmax}, it is clear that the curves peak at $1 + \sn{\rho_{\rm{S}}(\epsilon)^{\otimes n}}$.
\begin{figure}
    \centering
    \includegraphics[height=5cm]{figs/lc_strange.pdf}
    \caption{\textbf{Strange state Lorenz curves.} Lorenz curves of $\rho_{\rm{S}}(\epsilon)^{\otimes n}$ for $n=1,2,3,4$ in the $\sigma$--fragment where $\sigma = e^{-\beta H}/\Z_{\beta}$, with Hamiltonian spectrum $(0,1,2)$ over the computational basis and $\beta = 0.5$.
    Solid lines represent pure Strange states ($\epsilon = 0$); dashed lines represent noisy Strange states ($\epsilon = 0.1$).
    }
    \label{fig:lcs}
\end{figure}

>>>>

\newpage
\section{Phase space representations of quantum states}
\label{sec:ps}

Central to our construction is the representation of any quantum state or quantum operation on a system of dimension $d$ in terms of quasi-probability representations on a discrete phase space \ddd{[CITE the review on quasi-reps]}. This construction is a discrete version of Wigner representations in quantum optics.

We consider a $d$--dimensional quantum system with Hilbert space $\H_d$, and let $\{ |0\>, |1\>, \dots , |d-1\>\}$ denote the standard computational basis, defined over $\mathbb{Z}_d = \{ 0, 1, \dots,d-1 \}$. On this space, generalised Pauli matrices $X, Z$ can be defined by their respective roles as shift and phase operators, which act on the basis states as follows,
\begin{align}
    X \ket{k} &= \ket{k + 1} \label{eq:xpauli}\\
	Z \ket{k} &= \omega^k \ket{k}. \label{eq:zpauli}
\end{align}
Here $\omega \coloneqq e^{2\pi i/d}$ is the $d$-th root of unity and addition is taken modulo $d$. From these we can construct a phase space $\cal{P}_d = \mathbb{Z}_d \times \mathbb{Z}_d$ that provides a complete representation of the quantum system. Given a point $\bmx \coloneqq (x, p)$ we define a displacement operator, 
\begin{equation}\label{eq:ddef}
    D_{\bmx} \coloneqq \tau^{x p} X^{x} Z^{p},\ 
\end{equation}
where a phase factor $\tau \coloneqq -\omega^{1/2}$ ensures unitarity. We assume going forward that $d$ is an odd prime, however the case $d=2$ can be also handled, but with some additional technical caveats \ddd{[CITE]}. For a composite system with composite dimension $d = d_1 \dots d_n$ we can decompose the Hilbert space as $\cal{H}_d = \cal{H}_{d_1} \otimes \dots \otimes \cal{H}_{d_n}$, and then define displacement operators as
\begin{equation}\label{eq:composited}
    D_{\bmx} \coloneqq D_{(x_1, p_1)} \otimes \dots \otimes D_{(x_n, p_n)},
\end{equation}
where now we have
\begin{align*}
	\bmx \coloneqq (x_1, p_1, x_2, p_2, \dots, x_n, p_n) \in \cal{P}_{d_1} \times \dots \times \cal{P}_{d_n} \eqqcolon  \cal{P}_d,
\end{align*}
to denote the phase space point for the composite system. For simplicity going forward we assume $n$ copies of a $d$--dimensional system $d_1=d_2 = \cdots = d$, and therefore, we have that $\x \in \mathbb{Z}_d^{2n}$.

The displacement operators form the Heisenberg-Weyl group~\cite{Folland_1989, Bengtsson_2006} under matrix multiplication modulo phases,
\begin{equation}\label{eq:gp}
    {\rm{HW}}_d^n \coloneqq \{ \tau^k D_{\bmx}: k \in \mathbb{Z}_d, \bmx \in \cal{P}_d^n\}.
\end{equation}
The Clifford operations $ \cal{C}_d^n $ are then defined as the set of unitaries that normalise the Heisenberg-Weyl group \ddd{[CITE]}. We may define the pure stabiliser states as those states obtained by acting on $|0\>$ with Clifford unitaries. Finally, we define $\stab$ as the convex hull of all pure stabiliser states, namely all probabilistic mixtures of states of the form $U|0\>\<0|U^\dagger$ where $U$ is some Clifford unitary.

@@

In order to provide a complete decomposition of arbitrary quantum states and quantum operations we now define a complete basis of hermitian observables that behaves naturally under the action of the Clifford group. To every point $\x \in \P_d$ we define the phase-point operator $A_{\x}$ given by
\begin{align}
	A_{\bmx} \coloneqq \frac{1}{d} \sum_{\bmz \in \cal{P}_d} \omega^{\eta(\bmx, \bmz)} D_{\bmz}, \label{eq:ax}
\end{align}
where $\eta(\bmx, \bmz)$ is the symplectic inner product between any two points $\x,\z \in \P_d$, and is given explicitly by
\begin{equation}
	\eta(\bmx, \bmz) \coloneqq \bmz^T \begin{pmatrix}
		0  & \id \\ %\mathbbm{O}_n  & \id_n \\
		-\id & 0 \\ %-\id_n & \mathbbm{O}_n \\
	\end{pmatrix} \bmx,
\end{equation}
where $0, \id$ denote the $n\times n$ zero and identity matrices.

The phase-point operators form an orthogonal (with respect to the Hilbert-Schmidt inner product) operator basis as shown in~\cref{app:wigner}.
Therefore, any quantum state $\rho \in \cal{B}(\cal{H}_d)$ can be expressed as a linear combination of them,
\begin{equation}
    \rho = \sum_{\bmz \in \cal{P}_d} \W[\bmz]{\rho} A_{\bmz},
\end{equation}
where the coefficient vector $\W{\rho}$ is the Wigner distribution of state $\rho$,
\begin{equation}\label{eq:wstate}
    \W[\bmx]{\rho} \coloneqq \frac{1}{d}\tr[A_{\bmx} \rho].
\end{equation}

For any quantum state $\rho$, the Wigner distribution $W_\rho(\bmx)$ is readily seen to be a $d^2$-dimensional quasi-probability distribution over $\cal{P}_d$ (see~\cref{app:wigner} for details). More precisely, $W_\rho(\x)$ is a real-valued function on $\P_d$ with the property that $\sum_{\x} W_\rho(x) = 1$.  In~\cref{fig:wstate_examples}, we show Wigner distributions of different types of qutrit states.

\begin{figure}%
    \centering
    \subfigure[][]{%
    \label{fig:maxmix}%
    \includegraphics[height=2cm]{figs/maxmixed.pdf}
    %\caption{Maximally mixed state $\frac{1}{3}\id$}%
    }\hspace{8pt}%
    \subfigure[][]{%
    \label{fig:zero}%
    \includegraphics[height=2cm]{figs/zerostate.pdf}
    %\caption{Zero state $\ketbra{0}{0}$}%
    }\\
    \subfigure[][]{%
    \label{fig:bound}%
    \includegraphics[height=2cm]{figs/boundstate.pdf}
    %\caption{Bound state}%
    }\hspace{8pt}%
    \subfigure[][]{%
    \label{fig:strange}%
    \includegraphics[height=2cm]{figs/strangestate.pdf}
    %\caption{Strange state $\ketbra{S}{S}$}%
    }
    \caption{\textbf{Qutrit Wigner distributions of varying magic.} 
    \subref{fig:maxmix} Maximally mixed state $\frac{1}{3}\id$; \subref{fig:zero} Stabilizer zero state $\ketbra{0}{0}$; \subref{fig:bound} A non-stabiliser Wigner-positive state; \subref{fig:strange} Magic Strange state $\ket{{\rm{S}}} = \frac{1}{\sqrt{2}}(\ket{1} - \ket{2})$, coined in~\cite{cit:veitch2}.
    }%
    \label{fig:wstate_examples}
\end{figure}

\subsection{Magic theories for quantum computation}
\label{sec:mono}

The state-injection model for quantum computation \ddd{[CITE]} is a key avenue to realising a full-scale, fault-tolerant quantum computer. Within the approach Clifford unitaries can be done in a robust, fault-tolerant way. However, due to the Eastin-Knill theorem \ddd{[CITE]}, it is impossible to have a universal set of transversal gates and so while Clifford unitaries can be realised transversally one must find ways around the Eastin-Knill restriction. The state-injection approach supplements the Clifford gates with noisy quantum states, called `magic states', that are outside the set of stabliser states and, when incorporated into Clifford circuits, provides universality. The obstacle to this is that the states are invariably noisy and so protocols must be employed so as to purify many copies of the magic states and improve the overall performance of the induced quantum gates. A central question is then: given a number $n$ of noisy magic states how many purified magic states can we obtain within the fault-tolerant set of quantum operations?

To address such questions both concrete distillation protocols have been developed, and there has been analysis of bounds of distillation rates within natural theoretical frameworks. These frameworks provide theories of magic in which one views magic states as ``resource'' states with respect to a natural class of quantum operations that are consider cheap, or ``free'' \ddd{[CITE Gour resource review]}. One very natural class of free operations considered are those obtained from Clifford operations, measurements and the ability to discard quantum systems. However there are several other candidate frameworks.

For any theory of magic a natural route to bounding the distillation rates obtainable is through the concept of a \emph{magic monotone}. A magic monotone is a real-valued function of any quantum state $\M(\rho)$ that is monotonically non-increasing under the free operations of the magic theory. More precisely $\M(\sigma) \le \M(\rho)$ whenever it is possible to convert $\rho$ into $\sigma$ using at least one of the free operations available (e.g. Clifford operations).

One of the most fundamental and commonly used magic monotones is the \emph{mana} of a state~\cite{cit:veitch2}, defined as
\begin{equation}
    \mana{\rho} \coloneqq \ln{(2\hspace{1pt}\sn{\rho}+1)},
\end{equation}
where the \emph{sum-negativity}~\cite{cit:veitch2} is the sum of the negative components in $\W{\rho}$,
\begin{equation}
    \sn{\rho} \coloneqq \sum\limits_{\bmx: \W[\bmx]{\rho} < 0} \abs{\W[\bmx]{\rho}}.
\end{equation}

Mana is an additive\footnote{It satisfies the condition $\mana{\rho_1 \otimes \rho_2} = \mana{\rho_1} + \mana{\rho_2}$ which is practical in distillation settings.} magic monotone, so it provides an analytical, necessary condition for many-copy magic state interconvertibility.

A range of other monotones have been developed in the literature\cite{Wang_2018} \ddd{[CITE MORE]}, and each provides a means to upper bound distillation rates within a particular theory of magic. In this work we shall consider a broad framework for magic, with conditions that any theory of magic should satisfy. However, in contrast to prior works we shall not construct magic monotones, but instead develop a generalisation to resource monotones. The approach we take will also allow us to address how distillation rates are constrained by physical properties -- such as a system having biased-noise, some equilibrium structure or operations that manifest invariances, such as axial symmetry around the qubit $Z$-axis.

@@

In~\cref{fig:lcs}, we illustrate the Lorenz curve $\lc{\rho_{\rm{S}}}{\sigma}$\footnote{Hereinafter, we may omit obvious variable dependencies like $\epsilon$ and $n$ for clarity} of pure and noisy $n$--copy Strange states in some thermal $\sigma$--fragment.
Due to~\cref{lem:lcmax}, it is clear that the curves peak at $1 + \sn{\rho_{\rm{S}}(\epsilon)^{\otimes n}}$.
\begin{figure}
    \centering
    \includegraphics[height=5cm]{figs/lc_strange.pdf}
    \caption{\textbf{A `heretical' family of Lorenz curves.} Traditionally, Lorenz curves are monotone increasing cumulant functions that reach a maximum value of $1$. In contrast, Lorenz curves for magic states break through the value of $L(x)=1$ due to the presence of negativity in the associated quasi-probability distribution. The above family of curves correspond to multiple copies of noisy Strange states $\rho_{\rm{S}}(\epsilon)^{\otimes n}$ for $n=1,2,3,4$ within a particular stabilizer fragment. Solid lines represent pure Strange states, while dashed lines represent noisy Strange states with depolarising noise $\epsilon = 0.1$.
    }
    \label{fig:lcs}
\end{figure}
